---
title: "European Flights and Measles Incidence: Data Analysis"
author: "Maisha Adil"
date: "May 23, 2026"
output: html_document
---

## Part 0: Preliminaries
```{r, message=FALSE, warning=FALSE}
library(readxl) # Data is in .xlsx format
library(tidyverse)
library(visreg) # For partial effect visualization
library(glmnet)
library(gam) 
library(randomForest)
library(SuperLearner)

# Read in measles.xlsx from EpidemiologicalData > OutputData
measles <- read_excel("measles.xlsx")

# Read in flights_xxxx.xlsx from FlightPass > OutputData
flights_2019 <- read_excel("flights_2019.xlsx")
flights_2020 <- read_excel("flights_2020.xlsx")
flights_2022 <- read_excel("flights_2022.xlsx")
```


## Part 1: Flight Data Processing

Step 1: Clean the flight data 
```{r}
# Rows are origin country, columns are destination country
clean_flights <- function(flight_df, year) {
  # Drop the first 2 rows
  flight_df <- flight_df[-c(1, 2), ]
  
  # Drop the 3 identifier columns; keep only the passenger count matrix
  id_cols <- c("Country_name", "Alpha_3", "Alpha_2")
  passenger_matrix <- flight_df[, !(names(flight_df) %in% id_cols)]
  
  # Convert each column to numeric (lapply produces error for UK)
  for (j in 1:ncol(passenger_matrix)) {
    passenger_matrix[[j]] <- as.numeric(passenger_matrix[[j]])
  }
  
  # Total incoming passengers per destination = column sums
  # na.rm = TRUE treats missing route values as 0 (no recorded flights)
  incoming <- colSums(passenger_matrix, na.rm = TRUE)
  
  # Return a data frame with one row per country
  data.frame(
    RegionName = names(incoming),
    flights    = as.numeric(incoming),
    Year       = year
  )
}

# Apply the function to each year's flight matrix
flights_2019_clean <- clean_flights(flights_2019, 2019)
flights_2020_clean <- clean_flights(flights_2020, 2020)
flights_2022_clean <- clean_flights(flights_2022, 2022)
```


Step 2-3: Impute missing flight years and merge data
```{r}
# Step 2: Fill in missing flight years

# We only have observed flight data for 2019, 2020, and 2022
# The imputation plan is as follows:
# 2011-2018: use 2019 as a pre-COVID baseline
# 2019: observed
# 2020: observed (COVID-suppressed)
# 2021: average of 2020 and 2022
# 2022: observed
# 2023: carry 2022 forward

# Build a country x year grid (32 countries x 13 years = 416 rows)
all_countries <- sort(unique(measles$RegionName))
all_years     <- 2011:2023

flights_full  <- expand.grid(
  RegionName = all_countries,
  Year       = all_years,
  stringsAsFactors = FALSE
)
flights_full$flights <- NA

# For each year group, match each country name to the corresponding
# row in the observed data and pull the flights value at that position

# 2011-2019: use 2019 as a pre-COVID baseline
rows_2011_2019 <- flights_full$Year %in% 2011:2019
idx <- match(flights_full$RegionName[rows_2011_2019], flights_2019_clean$RegionName)
flights_full$flights[rows_2011_2019] <- flights_2019_clean$flights[idx]

# 2020: observed
rows_2020 <- flights_full$Year == 2020
idx <- match(flights_full$RegionName[rows_2020], flights_2020_clean$RegionName)
flights_full$flights[rows_2020] <- flights_2020_clean$flights[idx]

# 2021: interpolate as average of 2020 and 2022
rows_2021 <- flights_full$Year == 2021
idx_2020 <- match(flights_full$RegionName[rows_2021], flights_2020_clean$RegionName)
idx_2022 <- match(flights_full$RegionName[rows_2021], flights_2022_clean$RegionName)
flights_full$flights[rows_2021] <- (flights_2020_clean$flights[idx_2020] +
                                    flights_2022_clean$flights[idx_2022]) / 2

# 2022 and 2023: use 2022 observed value
rows_2022_2023 <- flights_full$Year %in% c(2022, 2023)
idx <- match(flights_full$RegionName[rows_2022_2023], flights_2022_clean$RegionName)
flights_full$flights[rows_2022_2023] <- flights_2022_clean$flights[idx]

# Check there are no missing flight values after fill-in
cat("Flight rows still missing after fill-in:", sum(is.na(flights_full$flights)), "\n")


# Step 3: Merge flights into the measles panel

# The measles data is monthly, whereas flight data is annual.
# Each year of flight data attaches to all 12 months of that country.

df <- merge(measles, flights_full, by = c("RegionName", "Year"), all.x = TRUE)
```


## Part 2: Measles Data Processing (Combined Dataset with Flights)


Step 1: Impute missing MCV2 values
```{r}
# MCV2 (second-dose vaccination coverage) varies by country-year and has 21 missing country-years (252 missing rows in the monthly panel).
# Two-stage imputation:
# Stage 1: linear interpolation within country (Finland, Italy, Luxembourg, which are missing only some years)
# Stage 2: regression imputation using MCV1 + Year (Ireland, which is missing ALL years and cannot be interpolated within-country)

# Work at the country-year level since MCV2 doesn't vary within a year
mcv2_data <- unique(df[, c("RegionName", "Year", "MCV1", "MCV2")])
mcv2_data <- mcv2_data[order(mcv2_data$RegionName, mcv2_data$Year), ]

# Flag every row that was originally missing (preserves imputation status even after filling in values)
mcv2_data$MCV2_imputed <- as.integer(is.na(mcv2_data$MCV2))


# Stage 1: Within-Country Linear Interpolation

# Base R's approx() does linear interpolation.
# rule = 2 means "for years outside the observed range, carry forward the nearest observed value"
# (constant fill at endpoints, rather than leaving them as NA).

for (ctry in unique(mcv2_data$RegionName)) {
  rows   <- which(mcv2_data$RegionName == ctry)
  values <- mcv2_data$MCV2[rows]
  years  <- mcv2_data$Year[rows]
  
  # Skip if all NA (move to Stage 2) or no NA (nothing to impute)
  if (all(is.na(values))) next
  if (!any(is.na(values))) next
  
  # Interpolate at all years using the non-missing values as anchors
  interp <- approx(
    x    = years[!is.na(values)],
    y    = values[!is.na(values)],
    xout = years,
    rule = 2
  )$y
  
  mcv2_data$MCV2[rows] <- interp
}


# Stage 2: Regression Imputation

# Fit MCV2 ~ MCV1 + Year on the country-years where MCV2 is now non-NA 
# (i.e. originally observed PLUS Stage-1 interpolated).
# Then predict MCV2 for the remaining missing rows (Ireland).
mcv2_data$Year <- as.integer(mcv2_data$Year)
mcv2_model <- lm(MCV2 ~ MCV1 + Year, data = mcv2_data)

cat("\nMCV2 imputation regression coefficients:\n")
print(round(summary(mcv2_model)$coefficients, 4))
cat("R-squared:", round(summary(mcv2_model)$r.squared, 3),
    "| Residual SD:", round(summary(mcv2_model)$sigma, 2), "pp\n")

# Predict for any rows still NA after Stage 1 (i.e. Ireland's 13 rows)
still_missing <- is.na(mcv2_data$MCV2)
mcv2_data$MCV2[still_missing] <- round(
  predict(mcv2_model, newdata = mcv2_data[still_missing, ]),
  1
)

# Merge imputed MCV2 and the imputation flag back into the main panel.
# First drop the original MCV2 column to avoid a name conflict.
df$MCV2 <- NULL
df <- merge(
  df,
  mcv2_data[, c("RegionName", "Year", "MCV2", "MCV2_imputed")],
  by   = c("RegionName", "Year"),
  all.x = TRUE
)
```


Step 2-3: Compute lagged case counts and reorganize file
```{r}
# Step 2: One-month lagged case counts

# This feature captures epidemic momentum. If cases were high last month, they are often elevated this month due to ongoing transmission.
# January 2011 has no prior month, so the first observation per country will have NA for counts_lag1.
# Keep these NAs and drop them only at the modeling step (and only for models that use this feature).

# Sort by country and time so the previous row really is the previous month
df <- df[order(df$RegionName, df$Time), ]
df$counts_lag1 <- NA

for (ctry in unique(df$RegionName)) {
  rows <- which(df$RegionName == ctry)
  # The lag at each row (except the first) is the previous row's cases
  df$counts_lag1[rows[-1]] <- df$Counts[rows[-length(rows)]]
}


# Step 3: Reorganize file with imputed data

# Extract the numeric month from "2011-01"-style strings
df$month <- as.integer(substr(df$Time, 6, 7))

# Rename for clarity
names(df)[names(df) == "RegionName"] <- "country"
names(df)[names(df) == "Counts"]     <- "cases"
names(df)[names(df) == "Pop_tot"]    <- "pop"
names(df)[names(df) == "Time"]       <- "month_str"

# Keep the columns we need in a sensible order
df <- df[, c("country", "Alpha3", "Year", "month", "month_str",
             "cases", "pop", "MCV1", "MCV2", "MCV2_imputed",
             "flights", "counts_lag1")]

# Confirm that no issues persist
cat(paste0("Final dataset: ", nrow(df), " rows x ", ncol(df), " columns\n"))
cat(paste0("MCV2 imputed: ", sum(mcv2_data$MCV2_imputed), " of 416 country-years (",
           sum(mcv2_data$MCV2_imputed) * 12, " monthly rows)\n"))

cat("\nColumns with any missing values:\n")
missing_counts <- sapply(df, function(x) sum(is.na(x)))
print(missing_counts[missing_counts > 0])
print(head(df))
```


## Part 3: Exploratory Data Analysis
```{r}
# Two-tone palette used across the section
burgundy  <- "#5f0a30"
green <- "#06402b"


# Step 1: Summary statistics and skewness

key_vars <- c("cases", "pop", "MCV1", "MCV2", "flights", "counts_lag1")
summary(df[, key_vars])

# Skewness numbers to motivate the log transform decision (Part 3+)
skewness_simple <- function(x) {
  x <- x[!is.na(x)]
  mean(((x - mean(x)) / sd(x))^3)
}
cat(paste0("Skewness (raw cases):      ", round(skewness_simple(df$cases), 2), "\n"))
cat(paste0("Skewness (log(cases + 1)): ", round(skewness_simple(log(df$cases + 1)), 2), "\n"))

# Zero-inflation
n_zero <- sum(df$cases == 0)
cat(paste0("Zero-case country-months: ", n_zero, " of ", nrow(df),
           " (", round(100 * n_zero / nrow(df), 1), "%)\n"))


# Step 2: Distribution of monthly cases

# Figure 1: Distribution of raw monthly measles cases
ggplot(df, aes(x = cases)) +
  geom_histogram(bins = 50, fill = burgundy, color = "white") +
  labs(title = "Figure 1. Distribution of Monthly Measles Cases (Raw)",
       x = "Monthly Cases Per country",
       y = "Number of Country-Months") +
  theme_minimal()

# Figure 2: Distribution of log(cases + 1)
ggplot(df, aes(x = log(cases + 1))) +
  geom_histogram(bins = 40, fill = burgundy, color = "white") +
  labs(title = "Figure 2. Distribution of log(Cases + 1) for Comparison",
       x = "log(Monthly Cases + 1)",
       y = "Number of Country-Months") +
  theme_minimal()


# Step 3: Distribution of vaccination coverage

# MCV1 and MCV2 vary only by country-year, so de-duplicate first
mcv_by_cy <- unique(df[, c("country", "Year", "MCV1", "MCV2")])

# Figure 3: Distribution of MCV1 first-dose coverage across country-years
ggplot(mcv_by_cy, aes(x = MCV1)) +
  geom_histogram(bins = 30, fill = green, color = "white") +
  geom_vline(xintercept = 95, color = "darkgrey", linetype = "dashed", linewidth = 0.6) +
  annotate("text", x = 94, y = Inf, label = "95% threshold",
           color = "darkgrey", hjust = 1, vjust = 1.5, size = 3) +
  labs(title = "Figure 3. Distribution of MCV1 Coverage (Country-Years)",
       x = "MCV1 First-Dose Coverage (%)",
       y = "Number of Country-Years") +
  theme_minimal()

# Figure 4: Distribution of MCV2 second-dose coverage across country-years
ggplot(mcv_by_cy, aes(x = MCV2)) +
  geom_histogram(bins = 30, fill = green, color = "white") +
  geom_vline(xintercept = 95, color = "darkgrey", linetype = "dashed", linewidth = 0.6) +
  annotate("text", x = 94, y = Inf, label = "95% threshold",
           color = "darkgrey", hjust = 1, vjust = 1.5, size = 3) +
  labs(title = "Figure 4. Distribution of MCV2 Coverage (Country-Years)",
       x = "MCV2 Second-Dose Coverage (%)",
       y = "Number of Country-Years") +
  theme_minimal()


# Step 4: Monthly cases by country, 2011-2023

df$date <- as.Date(paste0(df$month_str, "-01"))

# Figure 5: Monthly measles cases by country, 2011-2023
ggplot(df, aes(x = date, y = cases, group = country)) +
  geom_line(alpha = 0.4, color = burgundy, linewidth = 0.4) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(title = "Figure 5. Monthly Measles Cases by Country (n=32), 2011-2023",
       x = "Year",
       y = "Monthly Cases") +
  theme_minimal()
```


## Part 4: Candidate Model 1: Linear Regression

Shared Setup
```{r}
# Derived log columns
df$log_cases   <- log(df$cases + 1)
df$log_pop     <- log(df$pop)
df$log_flights <- log(df$flights + 1)

# Factor coercions
df$country <- as.factor(df$country)
df$month   <- as.factor(df$month)
df$Year    <- as.numeric(df$Year)

# Drop lag-NA rows (Jan 2011 per country); use df_model throughout Parts 4-9
df_model <- df[!is.na(df$counts_lag1), ]

# Temporal train/test split: train on 2011-2021, hold out 2022-2023 as test set.
# The approach preserves temporal ordering, so models never see future data during training.
df_train <- df_model[df_model$Year <= 2021, ]
df_test  <- df_model[df_model$Year >= 2022, ]

cat(paste0("Training set: ", nrow(df_train), " rows (", min(df_train$Year), "-", max(df_train$Year), ")\n"))
cat(paste0("Test set: ", nrow(df_test), " rows (", min(df_test$Year), "-", max(df_test$Year), ")\n"))
```


Linear Regression Predictor
```{r}
# Outcome: log(cases + 1)
# log_flights is included for a consistent feature set across all five models.
# Its coefficient is identified by COVID-era within-country variation (2020 dip,
# 2021 recovery), which confounds flight volume with broader pandemic disruptions.
# Do not interpret causally; see RF variable importance in Part 7.
lm_formula <- as.formula(
  "log_cases ~ MCV1 + MCV2 + log_pop + log_flights + counts_lag1 + Year + month + country"
)

# Fit on training set; evaluate on held-out test set
lm_fit_train <- lm(lm_formula, data = df_train)
lm_test_pred <- predict(lm_fit_train, newdata = df_test)
lm_test_mse  <- mean((df_test$log_cases - lm_test_pred)^2)
cat(paste0("Linear regression - Test MSE (2022-2023): ", round(lm_test_mse, 4), "\n"))

# Refit on full dataset for interpretation and plots
lm_fit_full <- lm(lm_formula, data = df_model)
summary(lm_fit_full)
```


Linear Regression Plots
```{r}
# Figure 6: Partial effect of MCV1 on log(cases + 1) - linear regression
visreg(lm_fit_full, "MCV1",
       ylab = "log(cases + 1)", xlab = "MCV1 coverage (%)",
       main = "Figure 6. Partial Effect of MCV1: Linear Regression",
       gg = FALSE)

# Figure 7: Partial effect of MCV2 on log(cases + 1) - linear regression
visreg(lm_fit_full, "MCV2",
       ylab = "log(cases + 1)", xlab = "MCV2 coverage (%)",
       main = "Figure 7. Partial Effect of MCV2: Linear Regression",
       gg = FALSE)

# Figure 8: Partial effect of prior month cases on log(cases + 1) - linear regression
visreg(lm_fit_full, "counts_lag1",
       ylab = "log(cases + 1)", xlab = "Prior month cases",
       main = "Figure 8. Partial Effect of Prior Month Cases: Linear Regression",
       gg = FALSE)

# Figure 9: Partial effect of log_flights on log(cases + 1) - linear regression
# Note: not causally interpretable
visreg(lm_fit_full, "log_flights",
       ylab = "log(cases + 1)", xlab = "log(incoming passengers + 1)",
       main = "Figure 9. Partial Effect of Flight Volume: Linear Regression",
       gg = FALSE)
```


## Part 5: Candidate Model 2: Regularized Methods (Lasso)

Lasso Predictor
```{r}
# glmnet requires a numeric matrix; expand factors with model.matrix()
# model.matrix() drops the intercept column; glmnet adds its own 
# Factor levels for country and month are identical across df_train and df_test
# (inherited from df_model), so both matrices have the same column set.

X_train_lasso <- model.matrix(~ MCV1 + MCV2 + log_pop + log_flights + counts_lag1 + Year + month + country,
                                data = df_train)[, -1]
y_train_lasso <- df_train$log_cases

X_test_lasso  <- model.matrix(~ MCV1 + MCV2 + log_pop + log_flights + counts_lag1 + Year + month + country,
                                data = df_test)[, -1]
y_test_lasso  <- df_test$log_cases

# Lambda selected via cv.glmnet on training data only (internal tuning, not evaluation)
set.seed(24)
cv_fit_train  <- cv.glmnet(X_train_lasso, y_train_lasso, alpha = 1, nfolds = 10)

# Predict on held-out test set using lambda.min
lasso_test_pred <- predict(cv_fit_train, newx = X_test_lasso, s = "lambda.min")
lasso_test_mse  <- mean((y_test_lasso - lasso_test_pred)^2)
cat(paste0("Lasso - Test MSE (2022-2023): ", round(lasso_test_mse, 4), "\n"))

# Refit on full dataset for coefficient interpretation and plots
X_full <- model.matrix(~ MCV1 + MCV2 + log_pop + log_flights + counts_lag1 + Year + month + country,
                         data = df_model)[, -1]
y_full         <- df_model$log_cases
cv_fit_full    <- cv.glmnet(X_full, y_full, alpha = 1, nfolds = 10)
lasso_fit_full <- glmnet(X_full, y_full, alpha = 1, lambda = cv_fit_full$lambda.min)

# Non-zero coefficients (excluding country and month dummies for clarity)
# coef() on a glmnet object returns a sparse matrix; as.matrix() converts it
# to a standard matrix so it can be put into a data frame below
coefs <- as.matrix(coef(lasso_fit_full))
coefs_df <- data.frame(
  variable = rownames(coefs),
  coefficient = as.numeric(coefs)
)
coefs_df <- coefs_df[coefs_df$coefficient != 0, ]
coefs_df <- coefs_df[!grepl("^country|^month", coefs_df$variable), ]
print(coefs_df[order(abs(coefs_df$coefficient), decreasing = TRUE), ])
```


Lasso Plots
```{r}
# visreg does not work directly with glmnet, so plot coefficients manually

# Extract all non-zero coefficients except intercept, country, and month dummies
coefs_plot <- coefs_df[coefs_df$variable != "(Intercept)", ]

# Sort once so barplot, names, and colors all use the same order
coefs_plot <- coefs_plot[order(coefs_plot$coefficient), ]

# Figure 10: Lasso non-zero coefficients (excluding country and month indicators)
ggplot(coefs_plot, aes(x = coefficient, y = reorder(variable, coefficient),
                        fill = coefficient > 0)) +
  geom_bar(stat = "identity") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_fill_manual(values = c("TRUE" = "#06402b", "FALSE" = "#5f0a30"),
                    guide = "none") +
  labs(title = "Figure 10. Lasso: Non-Zero Coefficients (Excluding Country/Month Indicators)",
       x = "Coefficient Value",
       y = NULL) +
  theme_minimal()
```


## Part 6: Candidate Model 3: Generalized Additive Models

GAM Predictor
```{r}
# MCV1 and MCV2 are entered as linear terms rather than smooth terms.
# An initial GAM with smooth terms performed worse than OLS during development.
# Country fixed effects absorb the between-country MCV1 variation that drives the visible 
# herd-immunity threshold in raw data; within-country changes over time are too small
# to reveal a threshold nonlinearity after conditioning on country. 

# log_flights included for a consistent feature set across all five models.
# As in Part 4, the coefficient is confounded by COVID-era disruptions and should not be interpreted causally. 
gam_formula <- as.formula(
  "log_cases ~ MCV1 + MCV2 + log_pop + log_flights + counts_lag1 + Year + month + country"
)

# Fit on training set; evaluate on held-out test set
gam_fit_train <- gam(gam_formula, data = df_train)
gam_test_pred <- predict(gam_fit_train, newdata = df_test)
gam_test_mse  <- mean((df_test$log_cases - gam_test_pred)^2)
cat(paste0("GAM - Test MSE (2022-2023): ", round(gam_test_mse, 4), "\n"))

# Refit on full dataset for interpretation and plots
gam_fit_full <- gam(gam_formula, data = df_model)
summary(gam_fit_full)
```


GAM Plots
```{r}
# Figure 11: Partial effect of MCV1 on log(cases + 1) - GAM
visreg(gam_fit_full, "MCV1",
       ylab = "log(cases + 1)", xlab = "MCV1 coverage (%)",
       main = "Figure 11. Partial Effect of MCV1: GAM",
       gg = FALSE)

# Figure 12: Partial effect of MCV2 on log(cases + 1) - GAM
visreg(gam_fit_full, "MCV2",
       ylab = "log(cases + 1)", xlab = "MCV2 coverage (%)",
       main = "Figure 12. Partial Effect of MCV2: GAM",
       gg = FALSE)

# Figure 13: Partial effect of prior month cases on log(cases + 1) - GAM
visreg(gam_fit_full, "counts_lag1",
       ylab = "log(cases + 1)", xlab = "Prior month cases",
       main = "Figure 13. Partial Effect of Prior Month Cases: GAM",
       gg = FALSE)

# Figure 14: Partial effect of log_flights on log(cases + 1) - GAM
# Note: not causally interpretable; see Section 3
visreg(gam_fit_full, "log_flights",
       ylab = "log(cases + 1)", xlab = "log(incoming passengers + 1)",
       main = "Figure 14. Partial Effect of Flight Volume: GAM",
       gg = FALSE)
```


## Part 7: Candidate Model 4: Random Forest

Training Dataset
```{r}
# Feature matrix and outcome; log_flights included since RF handles collinearity naturally
# and the training set now provides within-country variation in flight volume
features_rf <- c("MCV1", "MCV2", "log_pop", "log_flights", "counts_lag1", "Year", "month", "country")

# Feature matrices for training and test sets
X_train_rf <- df_train[, features_rf]
y_train_rf <- df_train$log_cases
X_test_rf  <- df_test[, features_rf]
y_test_rf  <- df_test$log_cases

# Tune mtry on training data only using OOB error
# Avoids optimistic bias from tuning on data used for evaluation
set.seed(24)
tune_train <- tuneRF(
  x          = X_train_rf,
  y          = y_train_rf,
  mtryStart  = floor(ncol(X_train_rf) / 3),
  ntreeTry   = 500,
  stepFactor = 1.5,
  improve    = 0.01,
  trace      = FALSE,
  plot       = FALSE
)
best_mtry_train <- tune_train[which.min(tune_train[, 2]), 1]
cat(paste0("Best mtry (training data): ", best_mtry_train, "\n"))

# Fit RF on training set; evaluate on held-out test set
set.seed(24)
rf_fit_train <- randomForest(
  x      = X_train_rf,
  y      = y_train_rf,
  ntree  = 500,
  mtry   = best_mtry_train
)

rf_test_pred <- predict(rf_fit_train, newdata = X_test_rf)
rf_test_mse  <- mean((y_test_rf - rf_test_pred)^2)
cat(paste0("Random Forest - Test MSE (2022-2023): ", round(rf_test_mse, 4), "\n"))
```


Full Dataset
```{r}
# Fit final RF on full dataset for variable importance
# X_rf and y_rf are also reused by Part 8's full-data SuperLearner fit
X_rf <- df_model[, features_rf]
y_rf <- df_model$log_cases

set.seed(24)
tune_full <- tuneRF(
  x          = X_rf,
  y          = y_rf,
  mtryStart  = floor(ncol(X_rf) / 3),
  ntreeTry   = 500,
  stepFactor = 1.5,
  improve    = 0.01,
  trace      = FALSE,
  plot       = FALSE
)
best_mtry_full <- tune_full[which.min(tune_full[, 2]), 1]
cat(paste0("Best mtry (full data): ", best_mtry_full, "\n"))

set.seed(24)
rf_fit_full <- randomForest(
  x          = X_rf,
  y          = y_rf,
  ntree      = 500,
  mtry       = best_mtry_full,
  importance = TRUE  # Required for varImpPlot
)

# Variable importance plot using permutation-based %IncMSE (type = 1)
# More reliable than node purity (type = 2) when predictors are correlated,
# as it directly measures prediction error increase when a feature is shuffled.
# Figure 15: Random forest feature importance (%IncMSE, permutation-based)
varImpPlot(rf_fit_full, type = 1, main = "Figure 15. Random Forest: Feature Importance (%IncMSE)")
```


## Part 8: Candidate Model 5: SuperLearner

SuperLearner Setup
```{r}
# Custom RF wrapper fixing mtry at the tuned value from Part 7
# Allows SL to compare default vs tuned RF as distinct candidates
SL.rf_mtry2 <- function(...) SL.randomForest(..., mtry = 2)

# Helper used by both the training wrapper and the predict method below
.gam_convert_factors <- function(df) {
  factor_cols <- sapply(df, is.factor)
  if (any(factor_cols))
    df[, factor_cols] <- lapply(df[, factor_cols], as.numeric)
  df
}

# Training wrapper: converts factor columns to integer level codes before SL.gam sees them.
# as.numeric() on a factor gives codes 1, 2, 3, ... for any factor regardless of label.
# Do NOT use as.numeric(as.character()): that produces NA for non-numeric labels like country names.
# The fit object is re-classed to "SL.gam_numeric" so that predict.SuperLearner
# dispatches to predict.SL.gam_numeric (below) rather than predict.SL.gam.
SL.gam_numeric <- function(Y, X, newX, family, obsWeights, ...) {
  result <- SL.gam(Y = Y, X = .gam_convert_factors(X),
                   newX = .gam_convert_factors(newX),
                   family = family, obsWeights = obsWeights, ...)
  class(result$fit) <- "SL.gam_numeric"
  result
}

# Predict method: mirrors the factor conversion so that post-training calls to
# predict(sl_fit_train, newdata = X_SL_test) work correctly.
# object$object is the stored gam model object (same structure as predict.SL.gam).
predict.SL.gam_numeric <- function(object, newdata, ...) {
  predict(object$object, newdata = .gam_convert_factors(newdata), type = "response")
}

# Candidate library
# create.Learner generates three glmnet wrappers across alpha values:
# alpha = 0 (ridge), alpha = 0.5 (elastic net), alpha = 1 (lasso)
enet <- create.Learner("SL.glmnet", detailed_names = TRUE,
                       tune = list(alpha = c(0, 0.5, 1)))

sl_lib <- c(
  "SL.mean",           # 1. Intercept-only baseline
  "SL.lm",             # 2. Linear regression (mirrors Part 4)
  enet$names,          # 3-5. Ridge, elastic net, lasso (mirrors Part 5)
  "SL.gam_numeric",    # 6. GAM with numeric month coercion (mirrors Part 6)
  "SL.randomForest",   # 7. Random forest with default mtry
  "SL.rf_mtry2"        # 8. Random forest with tuned mtry = 2 (mirrors Part 7)
)
```


SuperLearner Train/Test
```{r}
# Train/test matrices (reuse features_rf from Part 7)
X_SL_train <- df_train[, features_rf]
y_SL_train <- df_train$log_cases
X_SL_test  <- df_test[, features_rf]
y_SL_test  <- df_test$log_cases

# Fit SuperLearner on training data
# Inner 10-fold CV is used only for ensemble weight selection within the training set;
# it does not determine the reported test MSE
set.seed(24)
sl_fit_train <- SuperLearner(
  Y          = y_SL_train,
  X          = X_SL_train,
  SL.library = sl_lib,
  family     = gaussian(),
  method     = "method.NNLS",
  cvControl  = list(V = 10)
)

# Ensemble weights from the training fit
sl_fit_train

# Predict on held-out test set; compute test MSE
sl_test_pred <- predict(sl_fit_train, newdata = X_SL_test)$pred
sl_test_mse  <- mean((y_SL_test - sl_test_pred)^2)
cat(paste0("SuperLearner - Test MSE (2022-2023): ", round(sl_test_mse, 4), "\n"))
```


SuperLearner Full Dataset
```{r}
# Fit final SuperLearner on full dataset for ensemble weight interpretation
# X_rf and y_rf defined in Part 7 Full Dataset
set.seed(24)
sl_fit_full <- SuperLearner(
  Y          = y_rf,
  X          = X_rf,
  SL.library = sl_lib,
  family     = gaussian(),
  method     = "method.NNLS",
  cvControl  = list(V = 10)
)

# Print ensemble weights; candidates with weight = 0 were outperformed
# and not selected by the non-negative least squares optimizer
sl_fit_full
```


## Part 9: Model Evaluation

Model Comparison
```{r}
# Summary of 2022-2023 holdout test MSE across all five candidate models
model_comparison <- data.frame(
  Model     = c("Linear Regression", "Lasso", "GAM", "Random Forest", "SuperLearner"),
  Test_MSE  = c(lm_test_mse, lasso_test_mse, gam_test_mse, rf_test_mse, sl_test_mse),
  Test_RMSE = sqrt(c(lm_test_mse, lasso_test_mse, gam_test_mse, rf_test_mse, sl_test_mse))
)
model_comparison$Test_MSE  <- round(model_comparison$Test_MSE, 4)
model_comparison$Test_RMSE <- round(model_comparison$Test_RMSE, 4)

print(model_comparison)
```


Plot Setup
```{r}
# Use test set predictions already computed in Part 7
rf_eval <- data.frame(
  observed  = y_test_rf,
  predicted = rf_test_pred,
  country   = df_test$country,
  Year      = df_test$Year
)
```


Predicted vs Observed
```{r}
# Figure 16: Random forest predicted vs. observed log(cases + 1), 2022-2023 holdout
# 45-degree reference line indicates perfect prediction
# Points above the line are underpredicted; points below are overpredicted
ggplot(rf_eval, aes(x = predicted, y = observed)) +
  geom_point(alpha = 0.3, color = "#5f0a30", size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  labs(title = "Figure 16. Random Forest: Predicted vs Observed",
       subtitle = "2022-2023 Holdout Test Predictions",
       x = "Predicted log(cases + 1)",
       y = "Observed log(cases + 1)") +
  theme_minimal()
```


Bland-Altman Plot
```{r}
# Figure 17: Random forest Bland-Altman plot, 2022-2023 holdout
# x-axis: mean of predicted and observed (average magnitude)
# y-axis: predicted minus observed (signed error; positive = over-prediction)
# Mean difference (bias) and limits of agreement (+/- 1.96 SD) are shown

rf_eval$ba_mean <- (rf_eval$predicted + rf_eval$observed) / 2
rf_eval$ba_diff <- rf_eval$predicted - rf_eval$observed

bias <- mean(rf_eval$ba_diff)
sd_diff <- sd(rf_eval$ba_diff)
loa_upper <- bias + 1.96 * sd_diff
loa_lower <- bias - 1.96 * sd_diff

cat(paste0("Mean difference (bias): ", round(bias, 4), "\n"))
cat(paste0("Upper limit of agreement: ", round(loa_upper, 4), "\n"))
cat(paste0("Lower limit of agreement: ", round(loa_lower, 4), "\n"))

ggplot(rf_eval, aes(x = ba_mean, y = ba_diff)) +
  geom_point(alpha = 0.3, color = "#5f0a30", size = 0.8) +
  geom_hline(yintercept = bias, color = "#06402b", linetype = "solid", linewidth = 0.7) +
  geom_hline(yintercept = loa_upper, color = "grey40", linetype = "dashed", linewidth = 0.7) +
  geom_hline(yintercept = loa_lower, color = "grey40", linetype = "dashed", linewidth = 0.7) +
  annotate("text", x = max(rf_eval$ba_mean), y = bias,
           label = sprintf("Bias = %.3f", bias),
           hjust = 1, vjust = -0.5, size = 3, color = "#06402b") +
  annotate("text", x = max(rf_eval$ba_mean), y = loa_upper,
           label = sprintf("+1.96 SD = %.3f", loa_upper),
           hjust = 1, vjust = -0.5, size = 3, color = "grey40") +
  annotate("text", x = max(rf_eval$ba_mean), y = loa_lower,
           label = sprintf("-1.96 SD = %.3f", loa_lower),
           hjust = 1, vjust = 1.5, size = 3, color = "grey40") +
  labs(title = "Figure 17. Random Forest: Bland-Altman Plot",
       subtitle = "2022-2023 Holdout Test Predictions",
       x = "Mean of Predicted and Observed log(cases + 1)",
       y = "Predicted Minus Observed log(cases + 1)") +
  theme_minimal()
```

